Eco-evolutionary dynamics of finite populations from demographic first principles

### 2023-02-01 ind_files/03_BD_processes.tex
Minor changes in phrasing for Appendix B

Where for each $i$, we have defined the real-valued function $w_i(\mathbf{x}) \coloneqq b^{(\textrm{int})}_{i}(\mathbf{x}) - d^{(\textrm{int})}_i(\mathbf{x})$. Setting the mutation rate $\mu \to 0$ in equation \eqref{nD_det_limit_fitess_defn} makes it clear that the quantity $w_i(\mathbf{x})x_i$ describes the growth rate of type $i$ individuals in the population due to all processes other than mutation. The quantity $w_i(\mathbf{x})$ thus describes the per-capita growth rate of type $i$ individuals in a population $\mathbf{x}$, and is sometimes called the `Malthusian fitness' of type $i$. It is notable that the fitness of a type depends on the state of the population as a whole (\textit{i.e.} $\mathbf{x}$) and is thus frequency-dependent.\\

@@

where $B^{(1)}_t,B^{(2)}_t, \ldots, B^{(m)}_t$ are $m$ independent one-dimensional standard Brownian motion processes. The first term in this expression is the same as equation \eqref{nD_replicator_mutator} and describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation. As we can see, the replicator-mutator equation is generically satisfied for all birth-death processes that are in the (very general) functional forms \eqref{nD_functional_forms_for_replicator}. The second term is an additional directional force on population composition that acts exactly analogously to the replicator equation but in the opposite direction (A higher relative $\tau_i$ leads to a lower frequency), and furthermore, is particular to finite populations. This term describes a biasing effect due to differential turnover rates and can intuitively be understood in the following sense: If a type $i$ has a higher $\tau_i$, it experiences greater turnover due to a generally higher birth and death rate and thus greater demographic stochasticity. Types with a higher $\tau_i$ thus tend to be eliminated by a stochastic analog of selection because they experience more chance events (births and deaths) in a given time period (notice the minus sign before the second term in \eqref{nD_eqn_for_frequencies}). This effect is less visible if the total population size is higher because larger populations generally experience less stochasticity, and this is reflected in the $1/N$ factor in this term. This stochastic analog of selection for reduced turnover rates is the force responsible for the `reversal of the direction of deterministic selection' observed in previous studies \citep{constable_demographic_2016, mcleod_social_2019} and is an effect that is only seen in finite populations. Note that types that tend to increase the \emph{total} population size (such as altruists in evolutionary theory and mutualists in ecological communities) experience a smaller negative effect due to differential turnover rates than types that do not facilitate such an increase (such as cheaters and highly competitive species), which could explain why mutualists and cooperators routinely persist for long periods in finite population models despite extinction being predicted in infinite population models. The fact that the entire term is multiplied by $(KN(t))^{-1}$, however, suggests that the effect of this force is weak for medium to large populations, which explains why the aforementioned persistence of cooperators is often only observed in restrictive sounding conditions such as quasi-neutrality, rapid attraction to a slow manifold, or a weak selection limit. In all three of these cases, the first term on the RHS of \eqref{nD_eqn_for_frequencies} becomes identically 0 and therefore no longer contributes to the total dynamics, thus revealing the effects of differential turnover rate.\\
Finally, the last term describes non-directional stochastic effects due to fluctuations and has a `spreading effect'. To the highest order in $\sqrt{K}$, the effect of these fluctuations can be described through the weak noise approximation, as we explain below. However, the $1/K$ factor in the second term of \eqref{nD_eqn_for_frequencies} means that the effect of differential turnover rates disappears in this weak noise approximation  (and also in other such approximate treatments, such as the Chemical langevin).

>>>>

Where for each $i$, we have defined the real-valued function $w_i(\mathbf{x}) \coloneqq b^{(\textrm{int})}_{i}(\mathbf{x}) - d^{(\textrm{int})}_i(\mathbf{x})$. Setting the mutation rate $\mu \to 0$ in equation \eqref{nD_det_limit_fitess_defn} makes it clear that the quantity $w_i(\mathbf{x})x_i$ describes the growth rate of type $i$ individuals in the population due to all processes other than mutation. The quantity $w_i(\mathbf{x})$ thus describes the per-capita growth rate of type $i$ individuals in a population $\mathbf{x}$, and is sometimes called the `Malthusian fitness' of type $i$.  Ecologists often denote this quantity by the symbol $r_i$ and simply call it the (exponential) growth rate of type $i$, but we will stick to $w_i$ and `fitness' here. It is notable that the fitness of a type depends on the state of the population as a whole (\textit{i.e.} $\mathbf{x}$) and is thus frequency-dependent.\\

@@

where $B^{(1)}_t,B^{(2)}_t, \ldots, B^{(m)}_t$ are $m$ independent one-dimensional standard Brownian motion processes. The first term in this expression is the same as equation \eqref{nD_replicator_mutator} and describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation. We saw in the previous section that the replicator-mutator equation is generically satisfied in the infinite population limit for all birth-death processes that are in the (very general) functional forms \eqref{nD_functional_forms_for_replicator}, a result that is made explicit in \eqref{nD_eqn_for_frequencies} (just take $K \to \infty$ to convince yourself). The second term is an additional directional force on population composition that acts exactly analogously to the replicator equation but in the opposite direction (A higher relative $\tau_i$ leads to a lower frequency), and furthermore, is particular to finite populations. This term describes a biasing effect due to differential turnover rates and can intuitively be understood in the following sense: If a type $i$ has a higher $\tau_i$, it experiences greater turnover due to a generally higher birth and death rate and thus greater demographic stochasticity. Types with a higher $\tau_i$ thus tend to be eliminated by a stochastic analog of selection because they experience more chance events (births and deaths) in a given time period (notice the minus sign before the second term in \eqref{nD_eqn_for_frequencies}). This effect is less visible if the total population size is higher because larger populations generally experience less stochasticity, and this is reflected in the $1/N$ factor in this term. This stochastic analog of selection for reduced turnover rates is the force responsible for the `reversal of the direction of deterministic selection' observed in previous studies \citep{constable_demographic_2016, mcleod_social_2019} and is an effect that is only seen in finite populations. Note that types that tend to increase the \emph{total} population size (such as altruists in evolutionary theory and mutualists in ecological communities) experience a smaller negative effect due to differential turnover rates than types that do not facilitate such an increase (such as cheaters and highly competitive species), which could explain why mutualists and cooperators routinely persist for long periods in finite population models despite extinction being predicted in infinite population models. The fact that the entire term is multiplied by $(KN(t))^{-1}$, however, suggests that the effect of this force is weak for medium to large populations, which explains why the aforementioned persistence of cooperators is often only observed in restrictive sounding conditions such as quasi-neutrality, rapid attraction to a slow manifold, or a weak selection + weak mutation limit. In all three of these cases, the first term on the RHS of \eqref{nD_eqn_for_frequencies} becomes identically 0 and therefore no longer contributes to the total dynamics, thus uncovering the contributions of the second term.\\
Finally, the last term describes non-directional stochastic effects due to fluctuations and has a `spreading effect'. To the highest order in $\sqrt{K}$, the effect of these fluctuations can be described through the weak noise approximation, as we explain below. However, the $1/K$ factor in the second term of \eqref{nD_eqn_for_frequencies} means that the effect of differential turnover rates disappears in this weak noise regime  (and also in other such approximate treatments, such as the Chemical langevin).

### 2023-02-01 ind_files/99_Appendix_B.tex
Minor changes in phrasing for Appendix B

where $\boldsymbol{\mu}: \mathbb{R}^m \to \mathbb{R}^m$ is the `drift vector' and $\boldsymbol{\sigma}: \mathbb{R}^{m} \to \mathbb{R}^{m \times m}$ is the `diffusion matrix'. Let $f: \mathbb{R}^m \to \mathbb{R}$ be an arbitrary $C^2$ function. Then, It\^{o}'s lemma (\cite{oksendal_stochastic_1998}, Section 4.2) states that  the stochastic process $f(\mathbf{X}_t)$ must satisfy:

@@

Thus, $f_i(\mathbf{X}_t)$ gives us the frequency of type $i$ individuals when the population is described by the vector $\mathbf{X}_t$. We wish to use It\^{o}'s lemma \eqref{nD_Ito_lemma} to describe the change in this quantity. Let us, therefore, calculate each term in \eqref{nD_Ito_lemma}. We have the gradient:

@@

where we have defined $N = \sum_i x_i$ to represent the total population size to reduce clutter and used the fact that $\frac{\partial x_j}{\partial x_k} = \delta_{jk}$.
Similarly, for the Hessian, we have:

@@

Finally, using \eqref{nD_jacobian_for_ito} and denoting $d\mathbf{B}_t = [dB^{(1)}_t,dB^{(2)}_t, \ldots, dB^{(m)}_t]^{\mathrm{T}}$ where each $dB^{(j)}_t$ is an independent one dimensional Wiener process, we have:

@@

Thus, we see that the last term on the RHS of \eqref{nD_Ito_lemma} is given by

>>>>

where $\boldsymbol{\mu}: \mathbb{R}^m \to \mathbb{R}^m$ is the `drift vector' and $\boldsymbol{\sigma}: \mathbb{R}^{m} \to \mathbb{R}^{m \times m}$ is the `diffusion matrix'. Let $f: \mathbb{R}^m \to \mathbb{R}$ be an arbitrary $C^2(\mathbb{R}^m)$ function. Then, It\^{o}'s lemma (\cite{oksendal_stochastic_1998}, Section 4.2) states that  the stochastic process $f(\mathbf{X}_t)$ must satisfy:

@@

Thus, $f_i(\mathbf{X}_t)$ gives us the frequency of type $i$ individuals when the population is described by the vector $\mathbf{X}_t$. This function is obviously $C^2(\mathbb{R}^m)$, and we can thus use It\^{o}'s lemma \eqref{nD_Ito_lemma} to describe how it changes over time. The $j\textsuperscript{th}$ element of the gradient of $f_i$ is given by:

@@

where we have defined the total population size $N = \sum_i x_i$ and the frequency of the $i\textsuperscript{th}$ type $p_i = f_i(x)$ to reduce clutter, and used the fact that $\frac{\partial x_j}{\partial x_k} = \delta_{jk}$. The $jk\textsuperscript{th}$ element of the Hessian is given by:

@@

Finally, denoting $d\mathbf{B}_t = [dB^{(1)}_t,dB^{(2)}_t, \ldots, dB^{(m)}_t]^{\mathrm{T}}$ where each $dB^{(j)}_t$ is an independent one dimensional Wiener process, we have:

@@

Thus, using \eqref{nD_jacobian_for_ito}, we see that the last term on the RHS of \eqref{nD_Ito_lemma} is given by
